---
title: "Compare PCT statistics against funding data"
format: html
toc: true
number-sections: true
execute:
    echo: false
---


[Active travel LTA allocations](https://www.gov.uk/government/publications/active-travel-local-transport-authority-allocations/active-travel-local-transport-authority-allocations)

[RDEL and CDEL](https://www.gov.uk/government/publications/how-to-understand-public-sector-spending/how-to-understand-public-sector-spending#departmental-expenditure-limits-del)


```{r}
#| label: packages

# load packages

# install.packages("fuzzyjoin") # install fuzzyjoin package if not already installed

library(tidyverse) # for general functions
library(readr) # for reading csv files
library(sf) # for spatial data manipulation
library(dplyr) # for data manipulation
library(jsonlite) # for parsing JSON data
library(stringdist) # for string distance calculations
library(stringr) # for string manipulation

# library(tmap) # for mapping
library(ggplot2) # for plotting
# library(knitr) # for tables


```


# Data


## Results data

Results data obtained from results.csv file located in results folder in repo.

```{r}
#| label: import-results-data

# results.csv located in results folder

results <- read_csv("results/results.csv")

names(results) # check names in results

```


## LAD data search data

LAD names data was obtained from [uktransportauthorities](https://github.com/itsleeds/uktransportauthorities) repo.

Names of local authoriteis located in [transport_authorities_2026.geojson](https://github.com/itsleeds/uktransportauthorities/blob/main/transport_authorities_2026.geojson) file.

```{r}
#| label: import-lad-name-data

geojson_exists_locally <- file.exists("transport_authorities_2026.geojson")
if (geojson_exists_locally) {
    ltas <- sf::read_sf("transport_authorities_2026.geojson")
} else {
    ltas <- sf::read_sf("https://raw.githubusercontent.com/itsleeds/uktransportauthorities/refs/heads/main/transport_authorities_2026.geojson")
}

names(ltas) # check names in ltas

```


## Funding data

Funding for each local authority was obtained from [uktransportauthorities](https://github.com/itsleeds/uktransportauthorities/tree/main) repo. 

Data from 20_21 to 24_25 located in [atf_table_aggregated.csv](https://github.com/itsleeds/uktransportauthorities/blob/main/atf_table_aggregated.csv) file.

```{r}
#| label: import-funding-data

# Read data from atf_table_aggregated.csv file located in uktransportauthorities repo.
csv_exists_locally <- file.exists("atf_table_aggregated.csv")
if (!csv_exists_locally) {
    atf <- read_csv("https://raw.githubusercontent.com/itsleeds/uktransportauthorities/refs/heads/main/atf_table_aggregated.csv")
} else {
    atf <- read_csv("atf_table_aggregated.csv")
}

names(atf) # check names in atf

```

Data allocations from 2026 to 2030 located in [lad_active_travel_allocations.csv](https://github.com/itsleeds/uktransportauthorities/blob/main/lad_active_travel_allocations.csv) file.

```{r}
#| label: import-funding-data-future

# Read data from lad_active_travel_allocations.csv file located in uktransportauthorities repo.
lad_exists_locally <- file.exists("lad_active_travel_allocations.csv")
if (!lad_exists_locally) {
    atf_future <- read_csv("https://raw.githubusercontent.com/itsleeds/uk
transportauthorities/refs/heads/main/lad_active_travel_allocations.csv")
} else {
    atf_future <- read_csv("lad_active_travel_allocations.csv")
}

names(atf_future) # check names in atf_future

```

# Match LAD names

## Select relevant variables from results and ltas
Selected names from results and ltas to check against each other and match.

```{r}
#| label: select-names

# get list of unique la names
results_names <- results |>
    select(local_authority_name, combined_authority_name) |> # select name variables
    mutate(name = ifelse(
        !is.na(local_authority_name), 
        local_authority_name, 
        combined_authority_name
    )) |> # if lan not na, use it, otherwise can, then create new variable called "name"
    distinct(name) |> # unique values
    arrange(name) |> # sort alphabetically
    filter(!is.na(name)) # remove na values

# get list of names in ltas and remove geometry
ltas_names <- ltas |>
    st_drop_geometry() |> # remove geometry column
    select(name, lad_lookup) |> # select name and lad_lookup variables
    arrange(name) |> # sort alphabetically
    rename(name_lookup = lad_lookup) # rename lad_lookup to name_lookup

```


Unpack lad name from name_lookup column in ltas data to make it easier to match against results names.

```{r}
#| label: name-lookup-unpack

# unpack lad_lookup column in ltas_names data to make it easier to match against results names

ltas_lookup_names <- ltas_names |>
    mutate(name_lookup = map(
        name_lookup, 
        ~ fromJSON(.x)$lad_name
    )) |>
    select(name, name_lookup) |>
    unnest(name_lookup)

# if observation in name doesn't have exact match in name_lookup, add observation to name_lookup with same value as name

ltas_lookup_names <- bind_rows(
    ltas_lookup_names,
    ltas_lookup_names |>
        distinct(name) |>
        filter(!name %in% ltas_lookup_names$name_lookup) |>
        mutate(name_lookup = name)
)

# arrange alphabetically

ltas_lookup_names <- ltas_lookup_names |>
    arrange(name, name_lookup)

```


## Clean names for easier search and matching

Clean up results name to remove text such as "city" and "council" to make matching easier.

```{r}
#| label: results-names-clean

# remove unwanted text (not case sensitive)
results_names <- results_names |>
    mutate(name_clean = name |>
        str_remove_all("(?i)city") |>
        str_remove_all("(?i)councils") |>
        str_remove_all("(?i)council") |>
        str_remove_all("Borough") |>
        str_remove_all("(?i)district") |>
        str_remove_all("(?i)county") |>
        str_remove_all("(?i)of") |>
        str_remove_all("(?i)the") |>
        str_remove_all("(?i) and") |>
        str_remove_all("(?i)royal") |>
        str_remove_all("(?i)combined authority") |>
        str_remove_all("(?i)metropolitan") |>
        str_remove_all("(?i)(national guidance)") |>
        str_remove_all("(?i)&") |>
        str_remove_all("(?i),") |>
        str_remove_all("(?i)/") |>
        str_squish()
    )

```


## Add manual aliases for known mismatches

Add manual aliases for false positives and missing matches. This step may need to be updated and/or repeated after match-names step below.

```{r}
#| label: correct-ltas-lookup-names

# create tibble table of known aliases
manual_aliases_ltas <- tibble(
    name = c(
        "Kent",
        "West Yorkshire Combined Authority",
        "Bournemouth, Christchurch and Poole",
        "West of England Combined Authority",
        "Leicestershire",
        "Norfolk",
        "East Midlands Combined Authority",
        "North East Combined Authority",
        "East Riding of Yorkshire",
        "West Yorkshire Combined Authority",
        "Liverpool City Region Combined Authority",
        "Leicestershire",
        "Lancashire",
        "Leicestershire"
    ),
    name_lookup = c(
        "Faversham Town",
        "West Yorkshire",
        "BCP",
        "Bath North East Somerset Bristol South Gloucestershire North Somerset",
        "Blaby",
        "Broads Authority",
        "Derby Derbyshire Nottingham Nottinghamshire",
        "Durham",
        "Hull",
        "Leeds West Yorkshire",
        "Liverpool Region (LCRCA) in partnership with Halton Knowsley Liverpool Sefton St Helens Wirral",
        "North West Leicestershire",
        "Preston South Ribble Chorley",
        "Rutland"
    )
)

# append rows from manual_aliases to ltas_lookup_names
ltas_lookup_names <- bind_rows(ltas_lookup_names, manual_aliases_ltas) |>
    arrange(name, name_lookup)


```


Remove duplicates from ltas_lookup_names where there is a duplicate of name_lookup and name, keeping the first observation.

```{r}
#| label: remove-duplicates

ltas_lookup_names <- ltas_lookup_names |>
    distinct(name, name_lookup, .keep_all = TRUE) |>
    arrange(name, name_lookup)

```


## Match names between results and ltas

For each name obversation in results_names, search obversations in ltas_lookup_names$name_lookup to find the best match. If a best match is found, return the corresponding value in ltas_lookup_names$name. If there is no best match found, return NA. 

```{r}
#| label: match-names

# for each obversation in results_names$name_clean, search obversations in ltas_lookup_names$name_lookup to find the best match. If a best match is found, return the corresponding value in ltas_lookup_names$name. If there is no best match found, return NA.

results_names <- results_names |>
    mutate(
        match_index = amatch(
            name_clean, # search term
            ltas_lookup_names$name_lookup, # where to search
            method = "jw", # use Jaro-Winkler distance method
            maxDist = 0.25 # maximum distance for a match
        ),
        matched_name = ifelse(
            !is.na(match_index), # if match_index is not NA
            ltas_lookup_names$name_lookup[match_index], # return corresponding value in ltas_lookup_names$name_lookup
            NA_character_ # otherwise return NA
        ),
        ltas_name = ifelse(
            !is.na(match_index), # if match_index is not NA
            ltas_lookup_names$name[match_index], # return corresponding value in ltas_lookup_names$name
            NA_character_ # otherwise return NA
        )
    )

```


# atf name clean up

## Remove total row

CSV contains a total row which is not needed for analysis. Remove this row from atf data.

```{r}
#| label: remove-total-atf

# remove total row
atf <- atf |>
    filter(name != "Total")

```

## Update and add ATF names to match ltas_lookup_names

Change names in atf$name to match names in ltas_lookup_names$name. This step may need to be updated and/or repeated after check step below (check-ltas-against-atf).


```{r}
#| label: correct-atf-names

# change certain values in atf$name to match values in ltas_lookup_names$name.
atf_names <- atf |>
    mutate(name = case_when(
        name == "West Midlands ITA" ~ "West Midlands Combined Authority",
        name == "North East Joint Transport Committee" ~ "North East Combined Authority",
        TRUE ~ name
    ))

```


## Check ltas_lookup_names againts ATF LAD names

Check that all LAD names in ltas_lookup_names$name have an exact match to a LAD name in atf$name. If there are any mismatches, print them out.

```{r}
# check which names in ltas_lookup_names$name do not have an exact match to a name in atf_names$name

ltas_mismatches <- ltas_lookup_names |>
    filter(!name %in% atf_names$name) |>
    distinct(name) |>
    arrange(name)

# check which names in atf_names$name do not have an exact match to a name in ltas_lookup_names$name

atf_mismatches <- atf_names |>
    filter(!name %in% ltas_lookup_names$name) |>
    distinct(name) |>
    arrange(name)

```


The following mismatches are not in ltas_lookup_name but are in atf names - these are not an issue as they are older authorities which have been replaced by newer authorities which are present in both lists (these mismatches can be ignored for now):

- Northhamptonshire - replaced by North Northamptonshire
- Cumbria - replaced by Cumberland


The following mismatches in ltas_mismatches are not in atf_names but are present in the  atf_future (these mismatches can be ignored for now):

- Leicestershire
- North East Lincolnshire
- North Lincolnshire



# atf future name clean up

## Update atf future names

In atf_future need to:

- combine Derby, Derbyshire, Nottingham, Nottinghamshire, Leicester, Leicestershire, North East Lincolnshire, North Lincolnshire, Lincolnshire and Rutland into East Midlands Combined Authority.

This step may need to be updated and/or repeated after match check step below.

```{r}
#| label: correct-atf-future-names

# lower authorities in East Midlands Combined Authority
emca_la_list <- c(
    "Derby", 
    "Derbyshire", 
    "Nottingham", 
    "Nottinghamshire", 
    "Leicester", 
    "Leicestershire", 
    "North East Lincolnshire", 
    "North Lincolnshire", 
    "Lincolnshire", 
    "Rutland"
)

# filter values
emca_la <- atf_future |>
    filter(name %in% emca_la_list) |>
    summarise(across(where(is.numeric), sum, na.rm = TRUE)) |>
    mutate(
        name = "East Midlands Combined Authority", 
        name_atf = "East Midlands Combined County Authority"
    )

# bind emca_la to atf_future, removing lower authorities
atf_future_names <- atf_future |>
    filter(!name %in% emca_la_list) |>
    bind_rows(emca_la) |>
    arrange(name)

# remove duplicate rows
atf_future_names <- atf_future_names |>
    distinct(name, .keep_all = TRUE) |>
    arrange(name)

```

## check atf against atf future names

```{r}
#| label: check-ltas-against-atf

# check which names in ltas_lookup_names$name do not have an exact match to a name in atf_names$name

atf_future_mismatches_1 <- atf_future_names |>
    filter(!name %in% atf_names$name) |>
    distinct(name) |>
    arrange(name)

# check which names in atf_names$name do not have an exact match to a name in ltas_lookup_names$name

atf_future_mismatches_2 <- atf_names |>
    filter(!name %in% atf_future_names$name) |>
    distinct(name) |>
    arrange(name)

```

The following mismatches in atf_names are not in atf_future_names however can be ignored as they are older authorities that have been replaced:

- Cumbria
- Northamptionshire

Greater London is not in atf_future_names per future allocation figures being for authorities ["inside England **outside London**"](https://www.gov.uk/government/publications/active-travel-local-transport-authority-allocations/active-travel-local-transport-authority-allocations).

# Analysis

## Combine names into single table

```{r}
#| label: analysis-create-table

# combine name columns from results_names
analysis <- ltas_lookup_names |>
    select(name) |>
    distinct(name) |>
    arrange(name)


```

## Check names

Check to see what names (if any) are in data sources but not in analysis table

```{r}
#| label: check-names
# check what names in results$ltas_name are not in analysis$name
final_check <- results_names |>
    filter(!ltas_name %in% analysis$name) |>
    distinct(ltas_name) |>
    arrange(ltas_name) |>
    mutate(check = "results_names") |>
    rename(name = ltas_name)

# Create final_check_full table with names
final_check_full <- final_check


# check what names in atf_names$name are not in analysis$name

final_check <- atf_names |>
    filter(!name %in% analysis$name) |>
    distinct(name) |>
    arrange(name) |>
    mutate(check = "atf_names")

# append final_check to final_check_full
final_check_full <- bind_rows(final_check_full, final_check)

# check what names in atf_future_names$name are not in analysis$name

final_check <- atf_future_names |>
    filter(!name %in% analysis$name) |>
    distinct(name) |>
    arrange(name) |>
    mutate(check = "atf_future_names")

# append final_check to final_check_full
final_check_full <- bind_rows(final_check_full, final_check)

```


## Combine data

Match ltas_name from results_names to results

```{r}
#| label: results-ltas-name

results_analysis <- results |>
    left_join(
        results_names |>
            select(local_authority_name = name, ltas_name),
            by = "local_authority_name"
    )
```


Data wanted from results_analysis:

- number of observations for each ltas
- n_mentions_pct for each ltas
- for each ltas, pct_usage_depth of 1, 2, 3
- number of pct_desire_lines_used for each ltas


```{r}
#| label: analysis-add-results

# data from results_analysis

analysis_data <- analysis |>
    left_join(
        results_analysis |>
            mutate(pct_usage_depth = parse_number(pct_usage_depth)) |> # convert pct_usage_depth to numeric
            group_by(ltas_name) |>
            summarise(
                n_observations = n(), # number of observations for each ltas
                n_mentions_pct = sum(n_mentions_pct, na.rm = TRUE), # sum of n_mentions_pct for each ltas
                pct_usage_depth_1 = sum(pct_usage_depth == 1, na.rm = TRUE), # number of pct_usage_depth of 1 for each ltas
                pct_usage_depth_2 = sum(pct_usage_depth == 2, na.rm = TRUE), # number of pct_usage_depth of 2 for each ltas
                pct_usage_depth_3 = sum(pct_usage_depth == 3, na.rm = TRUE), # number of pct_usage_depth of 3 for each ltas
                n_pct_desire_lines_used = sum(pct_desire_lines_used, na.rm = TRUE) # number of pct_desire_lines_used for each ltas
            ),
            by = c("name" = "ltas_name")
    )

```


Data wanted from atf_names:

- atf20_21
- atf21_22
- atf22_23
- atf23_24
- atf24_25

Data wanted from atf_future_names:

- cdel_per_year_4yr
- rdel_per_year_3yr
- total_over_4yr


```{r}
#| label: analysis-add-atf

# add atf data. If no match, make value NA

analysis_data <- analysis_data |>
    left_join(
        atf_names |>
            select(name, atf20_21, atf21_22, atf22_23, atf23_24, atf24_25),
            by = "name"
    ) 

# add atf_future data

analysis_data <- analysis_data |>
    left_join(
        atf_future_names |>
            select(name, cdel_per_year_4yr, rdel_per_year_3yr, total_over_4yr),
            by = "name"
    )

# Clean funding columns: convert Invalid Number variants to NA, then parse numeric.
analysis_data <- analysis_data |>
    mutate(across(
        c(
            atf20_21,
            atf21_22,
            atf22_23,
            atf23_24,
            atf24_25,
            cdel_per_year_4yr,
            rdel_per_year_3yr,
            total_over_4yr
        ),
        ~ {
            x <- as.character(.x)
            x <- stringr::str_squish(x)
            x[stringr::str_detect(x, stringr::regex("^invalid\\s*number$", ignore_case = TRUE))] <- NA_character_
            readr::parse_number(x)
        }
    ))



```


# Compare PCT Usage Against Funding

## Build Comparable Metrics

```{r}
#| label: analysis-build-metrics

funding_5yr_cols <- c("atf20_21", "atf21_22", "atf22_23", "atf23_24", "atf24_25")

analysis_metrics <- analysis_data |>
    mutate(
        n_observations = replace_na(n_observations, 0),
        n_mentions_pct = replace_na(n_mentions_pct, 0),
        n_pct_desire_lines_used = replace_na(n_pct_desire_lines_used, 0),
        pct_usage_depth_1 = replace_na(pct_usage_depth_1, 0),
        pct_usage_depth_2 = replace_na(pct_usage_depth_2, 0),
        pct_usage_depth_3 = replace_na(pct_usage_depth_3, 0),
        n_depth_scored = pct_usage_depth_1 + pct_usage_depth_2 + pct_usage_depth_3,
        depth_score_mean = if_else(
            n_depth_scored > 0,
            (1 * pct_usage_depth_1 + 2 * pct_usage_depth_2 + 3 * pct_usage_depth_3) / n_depth_scored,
            NA_real_
        ),
        total_over_5yr = if_else(
            rowSums(!is.na(across(all_of(funding_5yr_cols)))) > 0,
            rowSums(across(all_of(funding_5yr_cols)), na.rm = TRUE),
            NA_real_
        ),
        pct_mention_rate = if_else(n_observations > 0, n_mentions_pct / n_observations, NA_real_),
        desire_lines_rate = if_else(n_observations > 0, n_pct_desire_lines_used / n_observations, NA_real_),
        mentions_per_million_funding = if_else(
            !is.na(total_over_5yr) & total_over_5yr > 0,
            n_mentions_pct / (total_over_5yr / 1e6),
            NA_real_
        )
    )

```

```{r}
#| label: analysis-rank-metrics

# variable to rank by
rank_variable <- "total_over_5yr" # variable to rank by

# ranked by rank_variable
ranked_analysis_metrics <- analysis_metrics |>
    select(
        name,
        n_observations,
        n_mentions_pct,
        pct_mention_rate,
        desire_lines_rate,
        depth_score_mean,
        total_over_5yr,
        total_over_4yr,
        mentions_per_million_funding
    ) |>
    arrange(desc(!!sym(rank_variable))) |>
    mutate(rank = row_number()) |>
    # Reduce number of decimal places to 1
    mutate(across(c(pct_mention_rate, desire_lines_rate, depth_score_mean, mentions_per_million_funding), ~ round(.x, 1)))
# knitr::kable(ranked_analysis_metrics, caption = paste("Ranked by", rank_variable))
DT::datatable(ranked_analysis_metrics, caption = paste("Ranked by", rank_variable), options = list(pageLength = 10, scrollX = TRUE))
```

https://assets.publishing.service.gov.uk/media/6939a16933c7ace9c4a41f8b/local-authority-active-travel-capability-ratings-2025.pdf
