---
title: "PCT / LCWIP Data Ingest & Funding Analysis (2026)"
subtitle: "REF 2029 impact case study — England proof-of-concept"
author: "Tom Auer"
date: last-modified
toc: true
number-sections: true
format:
  html: default
---

# Overview

This document reports the results of the **2026 re-run** of the data-ingest
process that collects evidence of the Propensity to Cycle Tool (PCT) and related
active-travel planning tools in English Local Cycling and Walking Infrastructure
Plans (LCWIPs) and related documents, integrated with official **Active Travel Fund (ATF)** allocation data.

The 2026 run is a **proof-of-concept for England** (the first stage before an
international roll-out). It improves on previous (~2024) work in four key ways:

1. **Distinction between LCWIP reports and related material** — each document is
   classified `LCWIP`, `LCWIP-related`, or `other`, a distinction the previous
   work did not make.
2. **Richer "how PCT was used" detail** — explicit scenario usage
   (`Go Dutch`, `Government Target`, `E-bike`, `Go Cambridge`), data sources,
   desire-line usage, and prioritisation-integration flags.
3. **Reproducible pipeline** — every web page / PDF is downloaded to
   `data-govuk-2026-raw` and a `.md` text extract written to
   `data-govuk-2026-md`; structured fields are extracted with a single,
   unlimited local engine (Ollama `gemma4-12b-hermes`) so the results are fully
   reproducible without paid API credits.
4. **Integration with Capital & Revenue Funding Data** — matching local authority
   entities to historical ATF funding (2020/21–2024/25) and multi-year future funding
   allocations (2026–2030) to evaluate how tool adoption aligns with infrastructure investment.

The headline dataset merges the fresh 2026 collection with the existing 94-entry
database (`LCWIP_database.json`), so the release shows **how the evidence base
has grown**.

# Ingest Methods

- **Discovery.** Seed URLs came from the existing 94-entry database, supplemented
  by web searches for 2024–2026 LCWIPs and related active-travel strategies, plus
  the Transport for the South East (TfSE) Regional Active Travel Strategy and
  Action Plan (RATSAP) Evidence Base (June 2025), which contains a 64-LCWIP
  status log for the South East.
- **Download.** Each candidate URL was fetched and stored as a raw file
  (`data-govuk-2026-raw`) with a text/markdown extract (`data-govuk-2026-md`).
- **Extraction.** A 21-field schema (`schema_2026.json`) was applied per document
  via a local LLM. Outputs are in `results/extracted/`.
- **Aggregation.** Extracted records were merged with the 94-entry database by
  URL/name and written to `results.json` and a flat `results.csv`.

# Corpus Composition

```{python}
#| label: corpus-composition
import pandas as pd
df = pd.read_csv("results/results_flat.csv")
print(f"Total merged records: {len(df)}")
print(f"  2026-only: {(df['_source_tag']=='2026').sum()}")
print(f"  from existing 94-DB: {(df['_source_tag']=='existing94').sum()}")
print()
print("Document type:")
print(df['doc_type'].fillna('(unknown)').value_counts())
```

# PCT Mentions

The central finding for the REF case study is the share of documents that
explicitly mention the Propensity to Cycle Tool.

```{python}
#| label: pct-mentions
import pandas as pd
df = pd.read_csv("results/results_flat.csv")
pct = df['mentions_pct']
n = pct.notna().sum()
mentioned = (pct == True).sum()
print(f"Documents with a usable mentions_pct flag: {n}")
print(f"Mention PCT: {mentioned} ({mentioned/n*100:.1f}%)")
print(f"Do not mention PCT: {(pct==False).sum()} ({ (pct==False).sum()/n*100:.1f}%)")
print(f"Unknown (no content / failed download): {pct.isna().sum()}")
```

## PCT Mentions by Document Type

```{python}
#| label: pct-by-doctype
import pandas as pd
df = pd.read_csv("results/results_flat.csv")
ct = pd.crosstab(df['doc_type'].fillna('unknown'), df['mentions_pct'].fillna('unknown'))
print(ct)
```

## Change Over Time

The chart below shows the evolution of LCWIP and active travel document publications over time (2018–2026), categorized by whether the report mentions the Propensity to Cycle Tool.

```{r}
#| label: fig-pct-over-time
#| fig-cap: "Publication of LCWIP and active-travel planning documents over time by PCT mention status."
#| warning: false
#| message: false
library(readr)
library(ggplot2)
library(dplyr)

csv_path <- "results/results_flat.csv"
if (!file.exists(csv_path)) csv_path <- "../results/results_flat.csv"

df <- read_csv(csv_path, show_col_types = FALSE) %>%
  mutate(year = readr::parse_number(as.character(date_published))) %>%
  filter(!is.na(year), year >= 2018, year <= 2026)

df_summary <- df %>%
  mutate(pct_label = if_else(mentions_pct, "Mentions PCT", "Does not mention PCT")) %>%
  group_by(year = factor(year), pct_label) %>%
  summarise(count = n(), .groups = "drop")

ggplot(df_summary, aes(x = year, y = count, fill = pct_label)) +
  geom_col(position = "stack", width = 0.6) +
  scale_fill_manual(values = c("Mentions PCT" = "#2b8cbe", "Does not mention PCT" = "#de2d26")) +
  labs(
    title = "LCWIP Document Publications and PCT Mentions Over Time",
    subtitle = "Annual count of published active travel strategy documents in England",
    x = "Publication Year",
    y = "Number of Documents",
    fill = "PCT Status"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "bottom",
    plot.title = element_text(face = "bold"),
    panel.grid.minor = element_blank()
  )
```

# Scenario Usage

Among documents that mention the PCT, which future scenarios are referenced?

```{python}
#| label: scenarios
import json, pandas as pd
from collections import Counter
df = pd.read_csv("results/results_flat.csv")
c = Counter()
for _, r in df.iterrows():
    if r.get('mentions_pct') is True or r.get('mentions_pct') == True:
        scs = r.get('pct_scenarios_used')
        if pd.notna(scs):
            if isinstance(scs, str):
                try: scs = json.loads(scs)
                except: pass
            if isinstance(scs, list):
                for s in scs:
                    c[str(s).strip()] += 1
sc = pd.Series(dict(c)).sort_values(ascending=False)
print("Scenario mentions (across PCT-mentioning docs):")
print(sc.to_string())
```

# Geographic Coverage

Coverage against higher-tier authorities in England gives a sense of how much of the country is represented in the corpus.

```{python}
#| label: geographic
import pandas as pd
df = pd.read_csv("results/results_flat.csv")
print("Records by region (England):")
print(df['region'].fillna('(not stated)').value_counts().to_string())
print()
print("Records with a combined authority named:")
print(df['combined_authority_name'].notna().sum())
```

# Cost and Network Length

Where stated, LCWIPs propose substantial networks and investment.

```{python}
#| label: cost-length
import pandas as pd
df = pd.read_csv("results/results_flat.csv")
cost = pd.to_numeric(df['total_cost_pounds'], errors='coerce')
km = pd.to_numeric(df['length_of_network_km'], errors='coerce')
print(f"Documents stating a total cost: {cost.notna().sum()}")
print(f"Total stated investment (£): {cost.sum():,.0f}")
print(f"Median stated investment (£): {cost.median():,.0f}")
print(f"Documents stating network length (km): {km.notna().sum()}")
print(f"Total stated network length (km): {km.sum():,.0f}")
```

# Illustrative Examples

A sample of LCWIPs that mention the PCT, with how it was used:

```{python}
#| label: examples
import pandas as pd
df = pd.read_csv("results/results_flat.csv")
ex = df[(df['mentions_pct']==True) & (df['doc_type']=='LCWIP')].head(10)
for _, r in ex.iterrows():
    print(f"### {r.get('report_name')} ({r.get('local_authority_name')})")
    print(f"- Date: {r.get('date_published')}")
    print(f"- Scenarios: {r.get('pct_scenarios_used')}")
    print(f"- How used: {str(r.get('how_pct_was_used') or '')[:300]}")
    print()
```

# Comparing PCT Evidence with Active Travel Funding

To evaluate the real-world policy and financial impact of the Propensity to Cycle Tool, we link the extracted LCWIP evidence with official UK Government funding datasets from the [Active Travel Local Transport Authority Allocations](https://www.gov.uk/government/publications/active-travel-local-transport-authority-allocations/active-travel-local-transport-authority-allocations), Departmental Expenditure Limits ([RDEL and CDEL](https://www.gov.uk/government/publications/how-to-understand-public-sector-spending/how-to-understand-public-sector-spending#departmental-expenditure-limits-del)), and [Active Travel England Capability Ratings 2025](https://assets.publishing.service.gov.uk/media/6939a16933c7ace9c4a41f8b/local-authority-active-travel-capability-ratings-2025.pdf).

```{r}
#| label: packages-funding
#| message: false
#| warning: false

library(tidyverse)
library(readr)
library(sf)
library(dplyr)
library(jsonlite)
library(stringdist)
library(stringr)
library(DT)
```

## Data Sources

```{r}
#| label: import-funding-sources
#| message: false
#| warning: false

# 1. Results data from extraction
results_file <- "results/results.csv"
if (!file.exists(results_file)) results_file <- "results/results_flat.csv"
if (!file.exists(results_file)) results_file <- "results_flat.csv"
results <- read_csv(results_file, show_col_types = FALSE)

# 2. LAD and Transport Authority geometry/names from uktransportauthorities
geojson_file <- "data/transport_authorities_2026.geojson"
if (!file.exists(geojson_file)) geojson_file <- "tda/transport_authorities_2026.geojson"
if (!file.exists(geojson_file)) geojson_file <- "transport_authorities_2026.geojson"

if (file.exists(geojson_file)) {
    ltas <- sf::read_sf(geojson_file)
} else {
    ltas <- sf::read_sf("https://raw.githubusercontent.com/itsleeds/uktransportauthorities/refs/heads/main/transport_authorities_2026.geojson")
}

# 3. Historical ATF Allocations (2020/21 - 2024/25)
atf_file <- "data/atf_table_aggregated.csv"
if (!file.exists(atf_file)) atf_file <- "tda/atf_table_aggregated.csv"
if (!file.exists(atf_file)) atf_file <- "scripts/atf_table_aggregated.csv"
if (!file.exists(atf_file)) atf_file <- "atf_table_aggregated.csv"

if (file.exists(atf_file)) {
    atf <- read_csv(atf_file, show_col_types = FALSE)
} else {
    atf <- read_csv("https://raw.githubusercontent.com/itsleeds/uktransportauthorities/refs/heads/main/atf_table_aggregated.csv", show_col_types = FALSE)
}

# 4. Future ATF Multi-Year Allocations (2026 - 2030)
future_file <- "data/lad_active_travel_allocations.csv"
if (!file.exists(future_file)) future_file <- "tda/lad_active_travel_allocations.csv"
if (!file.exists(future_file)) future_file <- "lad_active_travel_allocations.csv"

if (file.exists(future_file)) {
    atf_future <- read_csv(future_file, show_col_types = FALSE)
} else {
    atf_future <- read_csv("https://raw.githubusercontent.com/itsleeds/uktransportauthorities/refs/heads/main/lad_active_travel_allocations.csv", show_col_types = FALSE)
}
```

## Entity Reconciliation & Name Matching

To ensure accurate joins between LCWIP reports and funding allocations, local authority names are cleaned and matched using Jaro-Winkler string distance with manual alias rules for reconfigured unitary and combined authorities.

```{r}
#| label: match-lad-names
#| message: false
#| warning: false

# Extract unique authority names from extraction results
results_names <- results |>
    select(local_authority_name, combined_authority_name) |>
    mutate(name = ifelse(
        !is.na(local_authority_name), 
        local_authority_name, 
        combined_authority_name
    )) |>
    distinct(name) |>
    arrange(name) |>
    filter(!is.na(name))

# Extract names from LTAs and unpack JSON lookup
ltas_names <- ltas |>
    st_drop_geometry() |>
    select(name, lad_lookup) |>
    arrange(name) |>
    rename(name_lookup = lad_lookup)

ltas_lookup_names <- ltas_names |>
    mutate(name_lookup = map(
        name_lookup, 
        ~ fromJSON(.x)$lad_name
    )) |>
    select(name, name_lookup) |>
    unnest(name_lookup)

ltas_lookup_names <- bind_rows(
    ltas_lookup_names,
    ltas_lookup_names |>
        distinct(name) |>
        filter(!name %in% ltas_lookup_names$name_lookup) |>
        mutate(name_lookup = name)
)

# Standardise names by stripping council boilerplate
results_names <- results_names |>
    mutate(name_clean = name |>
        str_remove_all("(?i)city") |>
        str_remove_all("(?i)councils") |>
        str_remove_all("(?i)council") |>
        str_remove_all("Borough") |>
        str_remove_all("(?i)district") |>
        str_remove_all("(?i)county") |>
        str_remove_all("(?i)of") |>
        str_remove_all("(?i)the") |>
        str_remove_all("(?i) and") |>
        str_remove_all("(?i)royal") |>
        str_remove_all("(?i)combined authority") |>
        str_remove_all("(?i)metropolitan") |>
        str_remove_all("(?i)(national guidance)") |>
        str_remove_all("(?i)&") |>
        str_remove_all("(?i),") |>
        str_remove_all("(?i)/") |>
        str_squish()
    )

# Add manual aliases for known boundary reconfigurations
manual_aliases_ltas <- tibble(
    name = c(
        "Kent",
        "West Yorkshire Combined Authority",
        "Bournemouth, Christchurch and Poole",
        "West of England Combined Authority",
        "Leicestershire",
        "Norfolk",
        "East Midlands Combined Authority",
        "North East Combined Authority",
        "East Riding of Yorkshire",
        "West Yorkshire Combined Authority",
        "Liverpool City Region Combined Authority",
        "Leicestershire",
        "Lancashire",
        "Leicestershire"
    ),
    name_lookup = c(
        "Faversham Town",
        "West Yorkshire",
        "BCP",
        "Bath North East Somerset Bristol South Gloucestershire North Somerset",
        "Blaby",
        "Broads Authority",
        "Derby Derbyshire Nottingham Nottinghamshire",
        "Durham",
        "Hull",
        "Leeds West Yorkshire",
        "Liverpool Region (LCRCA) in partnership with Halton Knowsley Liverpool Sefton St Helens Wirral",
        "North West Leicestershire",
        "Preston South Ribble Chorley",
        "Rutland"
    )
)

ltas_lookup_names <- bind_rows(ltas_lookup_names, manual_aliases_ltas) |>
    distinct(name, name_lookup, .keep_all = TRUE) |>
    arrange(name, name_lookup)

# String distance fuzzy matching (Jaro-Winkler)
results_names <- results_names |>
    mutate(
        match_index = amatch(
            name_clean,
            ltas_lookup_names$name_lookup,
            method = "jw",
            maxDist = 0.25
        ),
        matched_name = ifelse(!is.na(match_index), ltas_lookup_names$name_lookup[match_index], NA_character_),
        ltas_name = ifelse(!is.na(match_index), ltas_lookup_names$name[match_index], NA_character_)
    )
```

## ATF Name Harmonisation

```{r}
#| label: clean-atf-names
#| message: false
#| warning: false

# Clean historical ATF table
atf <- atf |> filter(name != "Total")
atf_names <- atf |>
    mutate(name = case_when(
        name == "West Midlands ITA" ~ "West Midlands Combined Authority",
        name == "North East Joint Transport Committee" ~ "North East Combined Authority",
        TRUE ~ name
    ))

# Clean future allocations & aggregate East Midlands Combined County Authority
emca_la_list <- c(
    "Derby", "Derbyshire", "Nottingham", "Nottinghamshire", 
    "Leicester", "Leicestershire", "North East Lincolnshire", 
    "North Lincolnshire", "Lincolnshire", "Rutland"
)

emca_la <- atf_future |>
    filter(name %in% emca_la_list) |>
    summarise(across(where(is.numeric), sum, na.rm = TRUE)) |>
    mutate(
        name = "East Midlands Combined Authority", 
        name_atf = "East Midlands Combined County Authority"
    )

atf_future_names <- atf_future |>
    filter(!name %in% emca_la_list) |>
    bind_rows(emca_la) |>
    distinct(name, .keep_all = TRUE) |>
    arrange(name)
```

## Integrated Metric Construction

We aggregate document observations, depth-of-use scores, desire-line usage, and total funding amounts:

```{r}
#| label: build-integrated-dataset
#| message: false
#| warning: false

analysis <- ltas_lookup_names |>
    select(name) |>
    distinct(name) |>
    arrange(name)

results_analysis <- results |>
    left_join(
        results_names |> select(local_authority_name = name, ltas_name),
        by = "local_authority_name"
    )

analysis_data <- analysis |>
    left_join(
        results_analysis |>
            mutate(pct_usage_depth = parse_number(as.character(pct_usage_depth))) |>
            group_by(ltas_name) |>
            summarise(
                n_observations = n(),
                n_mentions_pct = sum(n_mentions_pct, na.rm = TRUE),
                pct_usage_depth_1 = sum(pct_usage_depth == 1, na.rm = TRUE),
                pct_usage_depth_2 = sum(pct_usage_depth == 2, na.rm = TRUE),
                pct_usage_depth_3 = sum(pct_usage_depth == 3, na.rm = TRUE),
                n_pct_desire_lines_used = sum(pct_desire_lines_used, na.rm = TRUE)
            ),
        by = c("name" = "ltas_name")
    ) |>
    left_join(
        atf_names |> select(name, atf20_21, atf21_22, atf22_23, atf23_24, atf24_25),
        by = "name"
    ) |>
    left_join(
        atf_future_names |> select(name, cdel_per_year_4yr, rdel_per_year_3yr, total_over_4yr),
        by = "name"
    ) |>
    mutate(across(
        c(atf20_21, atf21_22, atf22_23, atf23_24, atf24_25, cdel_per_year_4yr, rdel_per_year_3yr, total_over_4yr),
        ~ {
            x <- stringr::str_squish(as.character(.x))
            x[stringr::str_detect(x, stringr::regex("^invalid\\s*number$", ignore_case = TRUE))] <- NA_character_
            readr::parse_number(x)
        }
    ))

funding_5yr_cols <- c("atf20_21", "atf21_22", "atf22_23", "atf23_24", "atf24_25")

analysis_metrics <- analysis_data |>
    mutate(
        n_observations = replace_na(n_observations, 0),
        n_mentions_pct = replace_na(n_mentions_pct, 0),
        n_pct_desire_lines_used = replace_na(n_pct_desire_lines_used, 0),
        pct_usage_depth_1 = replace_na(pct_usage_depth_1, 0),
        pct_usage_depth_2 = replace_na(pct_usage_depth_2, 0),
        pct_usage_depth_3 = replace_na(pct_usage_depth_3, 0),
        n_depth_scored = pct_usage_depth_1 + pct_usage_depth_2 + pct_usage_depth_3,
        depth_score_mean = if_else(
            n_depth_scored > 0,
            (1 * pct_usage_depth_1 + 2 * pct_usage_depth_2 + 3 * pct_usage_depth_3) / n_depth_scored,
            NA_real_
        ),
        total_over_5yr = if_else(
            rowSums(!is.na(across(all_of(funding_5yr_cols)))) > 0,
            rowSums(across(all_of(funding_5yr_cols)), na.rm = TRUE),
            NA_real_
        ),
        pct_mention_rate = if_else(n_observations > 0, n_mentions_pct / n_observations, NA_real_),
        desire_lines_rate = if_else(n_observations > 0, n_pct_desire_lines_used / n_observations, NA_real_)
    )
```

## Ranked Metrics Table

The table below displays the resulting dataset ranked by 5-year historical ATF funding allocations (£), showing document counts, mention counts, mean depth score, and multi-year future allocations:

```{r}
#| label: ranked-metrics-table

ranked_analysis_metrics <- analysis_metrics |>
    select(
        name,
        n_observations,
        n_mentions_pct,
        pct_mention_rate,
        desire_lines_rate,
        depth_score_mean,
        total_over_5yr,
        total_over_4yr
    ) |>
    arrange(desc(total_over_5yr)) |>
    mutate(rank = row_number()) |>
    mutate(across(c(pct_mention_rate, desire_lines_rate, depth_score_mean), ~ round(.x, 1)))

DT::datatable(
    ranked_analysis_metrics,
    caption = "Local Transport Authorities ranked by 5-year ATF Allocations (£)",
    options = list(pageLength = 10, scrollX = TRUE)
)
```

```{r}
#| label: export-ranked-metrics
#| echo: false

# Save ranked metrics CSV for downstream impact summary
write_csv(ranked_analysis_metrics, "ranked_analysis_metrics.csv")
if (dir.exists("tda")) {
    write_csv(ranked_analysis_metrics, "tda/ranked_analysis_metrics.csv")
}
```

# Limitations & Next Steps

- **Geographic scope**: This 2026 run covers **England only**; international tools (NPT Scotland, CRUSE Ireland, biclaR Portugal, Norway PCT) are the focus of subsequent analysis stages.
- **Normalisation by authority scale**: Funding allocations are heavily influenced by local authority population and geographic scale; further work will normalise funding per capita.
- **Validation**: `doc_type` and free-text fields are model-extracted; key quotes and scoring should be spot-checked against source PDFs.
- **Reproducibility**: Extraction can be re-run locally with `scripts/run_extract_ollama.py` without external paid API credits.

