---
title: "PCT Tool and Funding Impact Analysis Summary"
author: "Tom Auer"
date: last-modified
format: html
toc: true
number-sections: true
execute:
    echo: false
    output: false
---

```{r}
#| label: load-packages

library(tidyverse) # for data manipulation and visualization
library(readr) # for reading csv files
library(scales) # for formatting numbers

```

# Metrics data

Analysis performed in `2026-analysis.qmd` generated ranked metrics for each transport authority region, saved as `ranked_analysis_metrics.csv`.

Load `ranked_analysis_metrics.csv` to view the ranked metrics data.

```{r}
#| label: load-ranked-metrics

metrics_path <- "ranked_analysis_metrics.csv"
if (!file.exists(metrics_path)) {
  metrics_path <- "tda/ranked_analysis_metrics.csv"
}

if (file.exists(metrics_path)) {
  ranked_metrics <- read_csv(metrics_path, show_col_types = FALSE)
} else {
  stop("File 'ranked_analysis_metrics.csv' does not exist. Please run the analysis in '2026-analysis.qmd' to generate the ranked metrics or check the file path.")
}

```

Rename columns for clarity

```{r}
#| label: rename-columns

# rename funding columns
ranked_metrics <- ranked_metrics |>
    rename(
        atf = total_over_5yr,
        atf_future = total_over_4yr
    )


```

Make atf and atf_future ranking columns

```{r}
#| label: atf-ranking-columns

# Create atf and atf_future ranking columns (high to low)
ranked_metrics <- ranked_metrics |>
    mutate(
        atf_rank = rank(-atf, ties.method = "min"),
        atf_future_rank = rank(-atf_future, ties.method = "min")
    )

# remove "rank" column if it exists
if ("rank" %in% colnames(ranked_metrics)) {
    ranked_metrics <- ranked_metrics |>
        select(-rank)
}

```


# Formatting (hide)

Make funding figures in format of £x,xxx,xxx

```{r}
#| label: format-funding-figures

# Format function
format_ranked_metrics <- function(df) {
    df |>
        mutate(
            atf = dollar( # format as £x,xxx,xxx
                atf, 
                prefix = "£", 
                big.mark = ",", 
                accuracy = 1
            ),
            atf_future = dollar( # format as £x,xxx,xxx
                atf_future, 
                prefix = "£", 
                big.mark = ",", 
                accuracy = 1
            )
        )
}
```


# Headline figures

Headline figures

- PCT adoption percentage
- Average pct mentions per document (of regions with at least 1 mention)
- Percentage of funding allocated to regions with mentions of PCT
- breakdown of depth score (e.g. x with 1, y with 2, z with 3, etc.)

Correlation
- Average funding per region with mentions of PCT vs no mentions
- caveat: authority size (larger authorities receive more funding more so based on size rather than PCT mentions)


```{r}
#| label: headline-figures

# count number of authorities with pct_mention_rate with Invalid Number
invalid_number_count <- ranked_metrics |>
    summarise(
        invalid_number_count = sum(is.na(pct_mention_rate))
    ) |>
    pull(invalid_number_count)

# PCT adoption percentage (count number of authorities with pct_mention_rate > 0 and not Invalid Number before dividing by total number of authorities)
pct_adoption_perc <- ranked_metrics |>
    summarise(
        pct_adoption_perc = sum(n_mentions_pct > 0) / (n() - invalid_number_count) * 100
    ) |>
    pull(pct_adoption_perc)

# Average pct mentions per document (of regions with at least 1 mention)
avg_pct_mentions_per_doc <- ranked_metrics |>
    filter(n_mentions_pct > 0) |>
    summarise(
        avg_pct_mentions_per_doc = sum(n_mentions_pct) / sum(n_observations)
    ) |>
    pull(avg_pct_mentions_per_doc)

# Percentage of funding allocated to regions with mentions of PCT (atf)
perc_atf_used_pct <-
    sum(ranked_metrics$atf[
        ranked_metrics$n_mentions_pct >0 & !is.na(ranked_metrics$atf)]
    ) / 
    sum(ranked_metrics$atf[
        !is.na(ranked_metrics$atf)]
    )

# Percentage of funding allocated to regions with mentions of PCT (atf_future)
perc_atf_future_used_pct <-
    sum(ranked_metrics$atf_future[
        ranked_metrics$n_mentions_pct >0 & !is.na(ranked_metrics$atf_future)]
    ) / 
    sum(ranked_metrics$atf_future[
        !is.na(ranked_metrics$atf_future)]
    )


# breakdown of depth score (e.g. x with >=1 and <2, y with >= 2 and <3, z with >=3, etc.)
depth_1_count <- ranked_metrics |>
    summarise(
        depth_1_count = sum(depth_score_mean >= 1 & depth_score_mean < 2, na.rm = TRUE)
    ) |>
    pull(depth_1_count)

depth_2_count <- ranked_metrics |>
    summarise(
        depth_2_count = sum(depth_score_mean >= 2 & depth_score_mean < 3, na.rm = TRUE)
    ) |>
    pull(depth_2_count)

depth_3_count <- ranked_metrics |>
    summarise(
        depth_3_count = sum(depth_score_mean >= 3, na.rm = TRUE)
    ) |>
    pull(depth_3_count)

# funding per depth score (atf)
funding_depth_1 <- ranked_metrics |>
    summarise(
        funding_depth_1 = sum(atf[depth_score_mean >= 1 & depth_score_mean < 2], na.rm = TRUE)
    ) |>
    pull(funding_depth_1)

funding_depth_2 <- ranked_metrics |>
    summarise(
        funding_depth_2 = sum(atf[depth_score_mean >= 2 & depth_score_mean < 3], na.rm = TRUE)
    ) |>
    pull(funding_depth_2)

funding_depth_3 <- ranked_metrics |>
    summarise(
        funding_depth_3 = sum(atf[depth_score_mean >= 3], na.rm = TRUE)
    ) |>
    pull(funding_depth_3)

funding_depth_2plus <- ranked_metrics |>
    summarise(
        funding_depth_2plus = sum(atf[depth_score_mean >= 2], na.rm = TRUE)
    ) |>
    pull(funding_depth_2plus)

# average funding per region with mentions of PCT vs no mentions (atf)
avg_atf_with_mentions <- ranked_metrics |>
    summarise(
        avg_fund_with_mentions = mean(atf[n_mentions_pct > 0], na.rm = TRUE)
    ) |>
    pull(avg_fund_with_mentions)

avg_atf_without_mentions <- ranked_metrics |>
    summarise(
        avg_fund_without_mentions = mean(atf[n_mentions_pct == 0], na.rm = TRUE)
    ) |>
    pull(avg_fund_without_mentions)

# average funding per region with mentions of PCT vs no mentions (atf_future)
avg_atf_future_with_mentions <- ranked_metrics |>
    summarise(
        avg_fund_with_mentions = mean(atf_future[n_mentions_pct > 0], na.rm = TRUE)
    ) |>
    pull(avg_fund_with_mentions)

avg_atf_future_without_mentions <- ranked_metrics |>
    summarise(
        avg_fund_without_mentions = mean(atf_future[n_mentions_pct == 0], na.rm = TRUE)
    ) |>
    pull(avg_fund_without_mentions)


```


```{r}
#| label: print-headline-figures
#| output: true

# Print headline figures with amounts formatted

print(paste0("PCT adoption percentage: ", round(pct_adoption_perc, 1), "%"))
print(paste0("Average PCT mentions per document (of regions with at least 1 mention): ", round(avg_pct_mentions_per_doc, 2)))
print(paste0("Percentage of funding allocated to regions with mentions of PCT (atf): ", round(perc_atf_used_pct * 100, 1), "%"))
print(paste0("Percentage of funding allocated to regions with mentions of PCT (atf_future): ", round(perc_atf_future_used_pct * 100, 1), "%"))
print(paste0("Regions with depth score of 1: ", depth_1_count))
print(paste0("Regions with depth score of 2: ", depth_2_count))
print(paste0("Regions with depth score of 3: ", depth_3_count))
print(paste0("Funding allocated to regions with depth score of 1 (atf): ", dollar(funding_depth_1, prefix = "£", big.mark = ",", accuracy = 1)))
print(paste0("Funding allocated to regions with depth score of 2 (atf): ", dollar(funding_depth_2, prefix = "£", big.mark = ",", accuracy = 1)))
print(paste0("Funding allocated to regions with depth score of 3 (atf): ", dollar(funding_depth_3, prefix = "£", big.mark = ",", accuracy = 1)))
print(paste0("Funding allocated to regions with depth score of 2 or more (atf): ", dollar(funding_depth_2plus, prefix = "£", big.mark = ",", accuracy = 1)))
print(paste0("Average funding per region with mentions of PCT (atf): ", dollar(avg_atf_with_mentions, prefix = "£", big.mark = ",", accuracy = 1)))
print(paste0("Average funding per region without mentions of PCT (atf): ", dollar(avg_atf_without_mentions, prefix = "£", big.mark = ",", accuracy = 1)))
print(paste0("Average funding per region with mentions of PCT (atf_future): ", dollar(avg_atf_future_with_mentions, prefix = "£", big.mark = ",", accuracy = 1)))
print(paste0("Average funding per region without mentions of PCT (atf_future): ", dollar(avg_atf_future_without_mentions, prefix = "£", big.mark = ",", accuracy = 1)))

```

```{r}
#| label: print top 10 atf and atf_future ranked regions
#| output: true

# Print top 10 regions by atf and atf_future using Knitr::kable for better formatting
top_10_atf <- ranked_metrics |>
    arrange(atf_rank) |>
    slice_head(n = 10) |>
    select(name, atf, atf_future, n_mentions_pct, pct_mention_rate, depth_score_mean)

top_10_atf_future <- ranked_metrics |>
    arrange(atf_future_rank) |>
    slice_head(n = 10) |>
    select(name, atf, atf_future, n_mentions_pct, pct_mention_rate, depth_score_mean)

knitr::kable(top_10_atf, caption = "Top 10 regions ranked by atf")
knitr::kable(top_10_atf_future, caption = "Top 10 regions ranked by atf_future")

```

# Summary


**91% of regions analysed use PCT** amount which referred to PCT an average of **12 times per document**.

**93% of current and 95% of future funding** is allocated to regions whose planning documents mention PCT.

Amoung regions for which depth of use can be assesed, **95% use PCT substantively rather than just simply referring to it** (average depth score of 2 or higher).

Regions mentioning PCT received substanitally more funding on average:

- **£9.4m compared to £4.5m** for current funding
- **£9.2m compared to £3.5m** for future funding

This difference in funding should be caveated by the fact that funding is also **influenced by authority size**. For current funding, Greater Manchester has the largest allocation of £63.2m but only 2 PCT mentiones whereas Surrey has 377 mentions with only £19.1m funding. Without considering authority size through factoring for population, the evidence suggests that **PCT has become embedded in the planning processes** through which the **vast majority** of active-travel investment is being delivered. This is strongly supported by the fact that **95% of regions demonstrated a substantive use of PCT**.

**Elevator pitch**: 

- **95%** of assessed regions demonstrated a **substantive use of PTC** in LCWIPs
- over **double** the average amount of **current funding** and almost **triple** the average amount of **future funding** was allocated to **regions mentioning PCT** compared to those that did not
- **93% of total current and 95% of total future funding** is allocated to regions whose planning documents mention PCT


# Further Steps

The analysis could be further strengthened by considering the following:

- **Authority size**: Consider normalising funding by authority size (e.g. population) to assess whether the difference in funding is due to PCT mentions or simply larger authorities receiving more funding
- **Depth of use**: Consider further analysis of the depth of use of PCT in planning documents to assess whether regions with more substantive use of PCT receive more funding than those with less substantive use
- **Temporal analysis**: Consider analysing the temporal aspect of PCT mentions in planning documents to assess whether regions that have been using PCT for longer periods of time receive more funding than those that have only recently started using PCT
- **Geographical analysis**: Consider analysing the geographical distribution of PCT mentions in planning documents to assess whether there are any regional patterns in the use of PCT and whether these patterns correlate with funding allocations

